An accelerated algorithm for finding efficient solutions in multiobjective problems with black-box multiextremal criteria\thanks{This work was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2022-08-18 2022/OPTIMA/paper.tex
fix the most part of the comments

Let $w_i(x), 1 \leq i \leq k,$ be real-valued functions depending on vector of arguments $x=(x_1,...,x_N)$  and defined over the hyperparallelepiped 
\begin{equation}\label{Q}
Q = \left\{x \in R^N : a_j \leq x_j \leq b_j, 1 \leq i \leq N \right\}
\end{equation}
in $N$-dimensional Euclidean space $R^N$. These functions reflect objectives of decision making and it is considered that the less the function value, the better the result of achieving this objective.

@@

It is known as a fundamental fact that $N$-dimensional hyperparallelepiped (\ref{Q}) and the interval $[0,1]$ of the real axis are the equipotent sets and the interval $[0,1]$ can be mapped onto the parallelepiped (\ref{Q}) unambiguously and continuously. Such mappings are called \textit{Peano-type curves} or \textit{evolvents}.

Let $x(t), \; t \in [0,1]$  be a Peano-type curve, and the function  $f(x)$ from (\ref{problem_f}) be continuous. As
\[
Q =\left\{x(t) : t \in [0,1]\right\},
\]
$f(x)$ and $x(t)$ are continuous, then
\[
\min_{x \in Q} f(x) = \min_{t \in [0,1]}f(x(t)),
\]
i.e., solving the multidimensional problem (\ref{problem_f}) can be reduced to the minimization of the one-dimensional function $f(x(t))$.

However, in the case when the function $f(x)$ is Lipschitzian this property is not valid for the reduced function. The function $f(x(t))$ will satisfy the H{\"o}lder condition
\begin{equation}\label{holder}
\left|f(x(t'))-f(x(t''))\right| \leq H \left|t'-t''\right|^{1/N}, \; t',t'' \in [0,1],
\end{equation}
with a H{\"o}lder constant $H>0$ \cite{Strongin2000,Sergeyev2013}.

If the function $f(x)$ satisfies the Lipschitz condition (\ref{lip_1}) then $H=2L\sqrt{N+3}$.
As a consequence, minimization of the function $f(x(t))$  requires the use of special algorithms oriented at functions with property (\ref{holder}).

@@

\begin{enumerate}
\item 
The points $t^1, \dots, t^n$ of completed trials are ordered in increasing order and renumbered by subscripts, i.e.,
\begin{equation}\label{eq:15}
t_1 = \alpha < t_1 < \dots < t_{n-1} < t_n = \beta.
\end{equation}
The values $v_s = \varphi(t_s)$ are juxtaposed to the points  $t_s$, $1 \leq s \leq n$.
\item 
A real values $R(q)$ are assigned to all the intervals $(t_{q-1}, t_q)$, $2 \leq q \leq n$, formed by neighboring points from (\ref{eq:15}). $R(q)$ is called \textit{the characteristic} of the interval $(t_{q-1}, t_q)$ and is computed in accordance with the following rule.

If number of completed trials is not exactly divisible by $T$, where \textit{global-to-local ratio} $T$ is a predefined number, then
\begin{equation}\label{eq:16}
R(q) = B(q),
\end{equation}
otherwise
\begin{equation}\label{eq:17}
R(q) = B(q)\left( \frac {\sqrt{(v_q - \varphi^*(n))(v_{q-1} - \varphi^*(n))}}{M} +(1.5)^{-\sigma}\right)^{-1}.
\end{equation}
Here
\begin{equation}\label{eq:18}
\begin{matrix}
B(q) = (t_s - t_{s-1})^{1/N} + \frac{(v_q - v_{q-1})^2} {m^2 (t_s - t_{s-1})^{1/N}} -2\frac{v_{q} + v_{q-1}}{m},\\
\varphi^*(n) = \min_{1\leq j \leq n} {\varphi(t^j)}
\end{matrix}
\end{equation}
is minimal computed value of the objective function $\varphi(t)$ at trial points,
\begin{equation}\label{eq:19}
M = \max_{2 \leq  s \leq n} { \frac{|v_s - v_{s-1}|}{(t_s - t_{s-1})^{1/N}} }
\end{equation}
is underestimation of the H{\"o}lder constant obtained according to the trial results,
\begin{equation}\label{eq:20}
m = 
\begin{cases}
\begin{matrix}
rM, & M>0, \\
1, & M= 0,
\end{matrix}
\end{cases}
\end{equation}
is improved estimation of the H{\"o}lder constant due to method's parameter $r>1$, $\sigma$ is some integer parameter that influences the speed of convergence to the global minimum.

\item
Among all intervals, the interval   with the highest characteristic is selected
\begin{equation}\label{eq:21}
R(p) = \max_{2 \leq s \leq n} {R(s)}.
\end{equation}

\item
The new $(n+1)$-th trial is performed at the point
\begin{equation}\label{eq:22}
t^{n+1} = \frac{t_p-t_{p-1}}{2} - sign(v_p-v_{p-1}) \left( \frac{|v_p-v_{p-1}|}{M} \right)^N \frac{1}{2r},
\end{equation}
the value $v^{n+1} = \varphi(t^{n+1})$ is computed and the iteration number $n$ is increased by 1.

\end{enumerate}

The method ALR described above is a modification of the basic information-statistical algorithm \cite{Strongin2000,Sergeyev2013} for minimization functions satisfying the H{\"o}lder condition. The modification consists in introducing new characteristics (\ref{eq:17}) after initial stage of the search to improve the method's speedup on account of increasing values of characteristics for intervals containing current record (\ref{eq:18}). 
Actually, the second term in (\ref{eq:17}) will have the largest value equal to $(1.5)^{\sigma}$ only for those intervals for which the equality $v_q = \varphi^*(n)$ or $v_{q-1} = \varphi^*(n)$ holds, i.e., for the intervals with the boundary points one of which coincides with the current record $\varphi^*(n)$.

Like its prototype, ARL can apply as the stopping condition the inequality
\begin{equation}\label{eq:23}
(t_p-t_{p-1})^{1/N} < \varepsilon,
\end{equation}
i.e., it completes the search when the length of the interval with maximal characteristic (\ref{eq:21}) becomes less than the predefined accuracy $\varepsilon > 0$.

@@

The complete convergence theory of the information-statistical algorithms is given in \cite{Strongin2000,Sergeyev2013}.

@@

In the course of optimization the algorithm builds the trial points $t^1,...,t^n$ which correspond to the points $x^1 = x(t^1),...,x^n=x(t^n)$ in the domain $Q$ . At these points the values $v^j = \Phi_\lambda(x^j), \; 1 \leq j \leq n,$ of the function $\Phi_\lambda(x)$ from (\ref{conv_problem}) are computed as values $\varphi(t^j)$ of the function $\varphi(t)$  minimized by ALR.

@@

In total, three groups of experiments have been performed. Each group contained a series of MCO problems of the same dimension and number of criteria. Since all GKLS functions have a global minimum value equal to $-1$, the partial  criteria $w_i(x), \; 1 \leq i \leq k,$  were increased by a positive constant to ensure the requirements (\ref{positive}).

In all experiments the domain
\[
Q = \left\{x \in R^N : -1 \leq x_j \leq -1, 1 \leq i \leq N \right\}
\]
was considered as the feasible one.

>>>>

Let $w_i(x), 1 \leq i \leq k,$ be real-valued functions depending on vector of arguments $x=(x_1,...,x_N)$  and defined over the hyperparallelepiped 
\begin{equation}\label{Q}
Q = \left\{x \in R^N : a_j \leq x_j \leq b_j, 1 \leq j \leq N \right\}
\end{equation}
in $N$-dimensional Euclidean space $R^N$. These functions reflect objectives of decision making and it is considered that the less the function value, the better the result of achieving this objective.

@@

It is known as a fundamental fact (see, e.g., \cite{Sagan1994}) that $N$-dimensional hyperparallelepiped (\ref{Q}) and the interval $[0,1]$ of the real axis are the equipotent sets and the interval $[0,1]$ can be mapped onto the parallelepiped (\ref{Q}) unambiguously and continuously. Such mappings are called \textit{Peano-type curves} or \textit{evolvents}.

Let $x(t), \; t \in [0,1],$  be a Peano-type curve, and the function  $f(x)$ from (\ref{problem_f}) be continuous. As
\[
Q =\left\{x(t) : t \in [0,1]\right\},
\]
$f(x)$ and $x(t)$ are continuous, then
\[
\min_{x \in Q} f(x) = \min_{t \in [0,1]}f(x(t)),
\]
i.e., solving the multidimensional problem (\ref{problem_f}) can be reduced to the minimization of the one-dimensional function $f(x(t))$.

%However, the function $f(x(t))$ will not meet the Lipschitz if $f(x)$ is Lipschitzian. It will satisfy the H{\"o}lder condition
However, in the case when the function $f(x)$ is Lipschitzian this property is not valid for the reduced function. The function $f(x(t))$ will satisfy the H{\"o}lder condition
\begin{equation}\label{holder}
\left|f(x(t'))-f(x(t''))\right| \leq H \left|t'-t''\right|^{1/N}, \; t',t'' \in [0,1],
\end{equation}
with a H{\"o}lder constant $H>0$.
And if the function $f(x)$ satisfies the Lipschitz condition (\ref{lip_1}) then $H=2L\sqrt{N+3}$ (see \cite{Sergeyev2013,Strongin2000}).
As a consequence, minimization of the function $f(x(t))$  requires the use of special algorithms oriented at functions with property (\ref{holder}).

@@

Step 1.
The points $t^1, \dots, t^n$ of completed trials are ordered in increasing order and renumbered by subscripts, i.e.,
\begin{equation}\label{eq:15}
t_1 = \alpha < t_1 < \dots < t_{n-1} < t_n = \beta.
\end{equation}
The values $v_s = \varphi(t_s)$ are juxtaposed to the points  $t_s$, $1 \leq s \leq n$.

Step 2. 
A real values $R(q)$ are assigned to all the intervals $(t_{q-1}, t_q)$, $2 \leq q \leq n$, formed by neighboring points from (\ref{eq:15}). $R(q)$ is called \textit{the characteristic} of the interval $(t_{q-1}, t_q)$ and is computed in accordance with the following rule.

If number of completed trials is not exactly divisible by $T$, where \textit{global-to-local ratio} $T$ is a predefined number, then
\begin{equation}\label{eq:16}
R(q) = B(q),
\end{equation}
otherwise
\begin{equation}\label{eq:17}
R(q) = B(q)\left( \frac {\sqrt{(v_q - \varphi^*(n))(v_{q-1} - \varphi^*(n))}}{M} +(1.5)^{-\sigma}\right)^{-1}.
\end{equation}
Here
\begin{gather}\label{eq:18}
B(q) = (t_s - t_{s-1})^{1/N} + \frac{(v_q - v_{q-1})^2} {m^2 (t_s - t_{s-1})^{1/N}} -2\frac{v_{q} + v_{q-1}}{m},\\
\varphi^*(n) = \min_{1\leq j \leq n} {\varphi(t^j)} \nonumber
\end{gather}
is minimal computed value of the objective function $\varphi(t)$ at trial points,
\begin{equation}\label{eq:19}
M = \max_{2 \leq  s \leq n} { \frac{|v_s - v_{s-1}|}{(t_s - t_{s-1})^{1/N}} }
\end{equation}
is underestimation of the H{\"o}lder constant obtained according to the trial results,
\begin{equation}\label{eq:20}
m = 
\begin{cases}
\begin{matrix}
rM, & M>0, \\
1, & M= 0,
\end{matrix}
\end{cases}
\end{equation}
is improved estimation of the H{\"o}lder constant due to method's parameter $r>1$, $\sigma$ is some integer parameter that influences the speed of convergence to the global minimum.

Step 3. 
Among all intervals, the interval   with the highest characteristic is selected
\begin{equation}\label{eq:21}
R(p) = \max_{2 \leq s \leq n} {R(s)}.
\end{equation}

Step 4.
The new $(n+1)$-th trial is performed at the point
\begin{equation}\label{eq:22}
t^{n+1} = \frac{t_p-t_{p-1}}{2} - sign(v_p-v_{p-1}) \left( \frac{|v_p-v_{p-1}|}{M} \right)^N \frac{1}{2r},
\end{equation}
the value $v^{n+1} = \varphi(t^{n+1})$ is computed and the iteration number $n$ is increased by 1.

The method ALR described above is a modification of the basic information-statistical algorithm \cite{Sergeyev2013,Strongin2000} for minimization functions satisfying the H{\"o}lder condition. The modification consists in introducing new characteristics (\ref{eq:17}) after initial stage of the search to improve the method's speedup on account of increasing values of characteristics for intervals containing current record (\ref{eq:18}). 
Actually, the second term in (\ref{eq:17}) will have the largest value equal to $(1.5)^{\sigma}$ only for those intervals for which the equality $v_q = \varphi^*(n)$ or $v_{q-1} = \varphi^*(n)$ holds, i.e., for the intervals with the boundary points one of which coincides with the current record $\varphi^*(n)$.

Like its prototype \cite{Sergeyev2013,Strongin2000}, ARL can apply as the stopping condition the inequality
\begin{equation}\label{eq:23}
(t_p-t_{p-1})^{1/N} < \varepsilon,
\end{equation}
i.e., it completes the search when the length of the interval with maximal characteristic (\ref{eq:21}) becomes less than the predefined accuracy $\varepsilon > 0$.

@@

The complete convergence theory of the information-statistical algorithms (including proof of the statements (\ref{eq:24}) and (\ref{eq:25})) is given in \cite{Sergeyev2013,Strongin2000}.

@@

In the course of optimization the algorithm generates the trial points $t^1,...,t^n$ which correspond to the points $x^1 = x(t^1),...,x^n=x(t^n)$ in the domain $Q$ . At these points the values $v^j = \Phi_\lambda(x^j), \; 1 \leq j \leq n,$ of the function $\Phi_\lambda(x)$ from (\ref{conv_problem}) are computed as values $\varphi(t^j)$ of the function $\varphi(t)$  minimized by ALR.

@@

In total, three groups of experiments have been performed. Each group contained a series of MCO problems of the same dimension and number of criteria. Since all GKLS functions have a global minimum value equal to $-1$, the partial  criteria $w_i(x), \; 1 \leq i \leq k,$  were increased by a positive constant to ensure the requirements (\ref{positive}).
In all experiments the domain $Q$ from (\ref{Q}) with $a_j=-1, \; b_j=1$ was considered as the feasible one.
